---
output: html_document
author: 
  - name: Claire Punturieri
    orcid: 0009-0003-7743-3736
    corresponding: true
    email: punturieri@wisc.edu
    url: https://arc.psych.wisc.edu/staff/punturieri-claire/
  - name: Christopher Janssen
    orcid: 0009-0001-2587-4367
    corresponding: true
    email: cjanssen3@wisc.edu
    url: https://github.com/christopher-janssen
editor_options: 
  chunk_output_type: console
---
# Visualizing geolocation data

```{r setup}
#| echo: false

# source
devtools::source_url("https://github.com/jjcurtin/lab_support/blob/main/fun_ml.R?raw=true")

# conflicts
options(conflicts.policy = "depends.ok")
tidymodels_conflictRules()
conflictRules("purrr", mask.ok = c("accumulate", "when"))
conflictRules("Matrix", mask.ok = c("expand", "pack", "unpack"))

# packages
library(tidyverse)
```

Visualizing geolocation data can be an excellent way to create fun visuals for talks and to make data checking easier. Below are several ways to display maps of geolocation data.

## Loading in and checking data.

For these tutorials we will be using data from a single individual collected during the Women's March on the Washington Mall using the Moves app. These data were pulled from a [publicly available online tutorial](https://quantdev.ssri.psu.edu/tutorials/intro-geospatial-data-and-maps-r).

```{r}
gps <- read_csv("data/moves_example_data.csv",
                show_col_types = FALSE) |>
  janitor::clean_names() |> 
  rename(lon = longitude,
         lat = latitude) |>
  glimpse()
```

## Creating static maps

Static maps can be created using the `ggmap` library. You will need to set an API key for Stadia Maps to access some of the available designs (specified using the maptype argument in `get_map()`).

```{r}
library(ggmap)
register_stadiamaps("d4bd71d7-556a-4627-9515-ef6e96823ce3")
```

Here is an example with points.
```{r}
cbbox <- make_bbox(lon = gps$lon, lat = gps$lat, f = .1)
sq_map <- get_map(location = cbbox, maptype = "stamen_terrain", source = "stadia")
    
ggmap(sq_map) + 
  geom_point(data = gps, aes(x = gps$lon, y = gps$lat), 
            size = 3, alpha = 0.6, color="darkred") +
  labs(x = " ", y = " ", title = "Static map (points)") +
  theme_minimal() +
  theme(legend.position = "none")
```

And an example with points that vary in size based on duration.
```{r}
cbbox <- make_bbox(lon = gps$lon, lat = gps$lat, f = .1)
sq_map <- get_map(location = cbbox, maptype = "stamen_terrain", source = "stadia")
    
ggmap(sq_map) + 
  geom_point(data = gps, aes(x = gps$lon, y = gps$lat, size = duration), 
            alpha = 0.6, color="darkred") +
  labs(x = " ", y = " ", title = "Static map (points based on dur.)") +
  theme_minimal() +
  theme(legend.position = "none")
```

Here is an example with connected lines if you were interested in looking at the travel path that an individual took using a `geom_path()` layer.
```{r}
cbbox <- make_bbox(lon = gps$lon, lat = gps$lat, f = .1)
sq_map <- get_map(location = cbbox, maptype = "stamen_terrain", source = "stadia")
    
ggmap(sq_map) + 
  geom_path(data = gps, aes(x = gps$lon, y = gps$lat), 
            linewidth = 1, lineend = "round", color="darkred") +
  labs(x = " ", y = " ", title = "Static map (path traveled)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Creating dynamic maps

Dynamic maps can be helpful visual tools for QC in addition to giving your audience something interesting to look at.

### Animated maps

We can create animated maps using moveVis, with help from move2 and raster.

```{r}
# devtools::install_github("16EAGLE/moveVis")
library(moveVis)
library(raster, exclude = c("select"))
library(move2)
library(magick)
```

You will need to first filter out any locations with less than 2 points (there is one such point in our location data). A later function will also be looking for an identification column, which will likely be present for any research participant data (but isn't in these mock data, so we'll add it in).
```{r}
gps_filtered <- gps |> 
  group_by(name) |> 
  filter(n() >= 2) |> 
  ungroup() |> 
  mutate(subid = 1)
```

Next, we need to convert the GPS data into a move object.
```{r}
gps_moves <- df2move(gps_filtered,
        proj = "EPSG:4326",  # specifies the coordinate reference system
        x = "lon", y = "lat",
        time = "start", # in other datasets, this might be dttm_obs
        track_id = "subid") # your subject identifier goes here
```

Then, we take our move object and interpolate it to be at regular time intervals.
```{r}
m <- align_move(gps_moves,
                res = 5, unit = "mins") # data will be aligned to every 5 minutes
```

Now we can overlay these movement patterns onto a map.
```{r}
frames <- frames_spatial(m, path_colours = "darkblue",
                         map_service = "osm_stamen", map_type = "terrain",
                         map_token = "d4bd71d7-556a-4627-9515-ef6e96823ce3", # API key
                         alpha = 0.5, path_legend = FALSE) |>  
  add_labels(x = "Longitude", y = "Latitude") |>
  add_northarrow() |>  
  add_scalebar() |> 
  add_timestamps(type = "label") |>  
  add_progress()
```

In general, it's a good idea to look at some frames to make sure the map looks correct. You can index into your frames object using the following code.
```{r}
frames[[10]]
```

Once you've checked a few frames and it looks like everything is rendering as it should, you can save out the animation as a gif.
```{r}
outfile <- "data/moveVis.gif"

if (!file.exists(outfile)) {
  animate_frames(frames, out_file = outfile)
} else {
  message("File already exists, skipping animation rendering.")
}
```

image_read() from the magick package can be used to display gifs (also works in your Viewer in RStudio!).
```{r}
img <- image_read("data/moveVis.gif")
print(img)
```

### Interactive maps

Interactive maps using Leaflet provide an excellent way to explore GPS data with some new functionality.

```{r}
library(leaflet)
library(htmlwidgets)
```

#### Basic Interactive Leaflet Map

Here's a basic interactive map with clickable points:

```{r}
leaflet(gps) |>
  addTiles() |>  # adds the default openstreetmap tiles
  addCircleMarkers(
    lng = ~lon,  # longitude column from data
    lat = ~lat,  # latitude column from data
    radius = 6,  # size of circles in pixels
    color = "#000",  # border color
    fillColor = "#C5050C",  # fill color (go badgers)
    fillOpacity = 0.7,  # transparency (0 = transparent, 1 = opaque)
    stroke = TRUE,  # whether to draw border
    weight = 2,  # border thickness
    popup = ~paste0(  # html popup when clicking points
      "<strong>", name, "</strong><br>",
      "Date: ", date, "<br>",
      "Start: ", format(as.POSIXct(start), "%H:%M"), "<br>",
      "Duration: ", round(duration/60, 1), " minutes"
    )
  ) |>
  # center map on average coordinates
  setView(lng = mean(gps$lon), lat = mean(gps$lat), zoom = 13)
```

#### Map with Duration-based Sizing

We can also vary the point size based on duration to show how long the person spent at each location:

```{r}
leaflet(gps) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~lon, 
    lat = ~lat,
    # scale radius by duration with minimum size of 3 pixels
    radius = ~pmax(3, duration/200),  
    color = "#000",  # black border
    fillColor = "#C5050C",  # red fill (go badgers)
    fillOpacity = 0.6,
    stroke = TRUE,
    weight = 1,  # thinner border than previous example
    popup = ~paste0(
      "<strong>", name, "</strong><br>",
      "Duration: ", round(duration/60, 1), " minutes<br>",
      "Time: ", format(as.POSIXct(start), "%H:%M")
    )
  ) |>
  setView(lng = mean(gps$lon), lat = mean(gps$lat), zoom = 13)
```

#### Map with Movement Path

For showing movement patterns, we can add lines connecting the locations in chronological order:

```{r}
# first sort data by time to ensure correct path order
gps_ordered <- gps |> arrange(as.POSIXct(start))

leaflet(gps_ordered) |>
  addTiles() |>
  # add polylines !first! so they appear under the points
  addPolylines(
    lng = ~lon,
    lat = ~lat,
    color = "#C5050C",  # go badgers
    weight = 3,  # line thickness
    opacity = 0.7  # line transparency
  ) |>
  # add circle markers on top of the lines
  addCircleMarkers(
    lng = ~lon, 
    lat = ~lat,
    radius = 5,
    color = "#000",  # black border
    fillColor = "#FFF", # white fill (go badgers)
    fillOpacity = 0.8,
    weight = 2,
    popup = ~paste0(
      "<strong>", name, "</strong><br>",
      "Time: ", format(as.POSIXct(start), "%H:%M"), "<br>",
      "Duration: ", round(duration/60, 1), " minutes"
    )
  ) |>
  setView(lng = mean(gps$lon), lat = mean(gps$lat), zoom = 13)
```

#### Map with Color-coding

We can also add color-coding to better visualize patterns in the data:

```{r}
# example with color-coded duration using viridis palette
gps_chron <- gps |>
  mutate(
    start_time = as.POSIXct(start),
    duration_min = round(duration / 60, 1)
  ) |>
  arrange(start_time)

# create color palette for duration
pal <- colorNumeric("viridis", domain = gps_chron$duration)

leaflet(gps_chron) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~lon, 
    lat = ~lat,
    radius = 6,
    color = "#000",
    fillColor = ~pal(duration),  # color by duration
    fillOpacity = 0.7,
    weight = 1,
    popup = ~paste0(
      "<strong>", name, "</strong><br>",
      "Duration: ", duration_min, " minutes<br>",
      "Time: ", format(start_time, "%H:%M")
    )
  ) |>
  addLegend(
    pal = pal, 
    values = ~duration,
    title = "Duration (sec)",
    position = "bottomright"
  ) |>
  setView(lng = mean(gps_chron$lon), lat = mean(gps_chron$lat), zoom = 13)
```

#### Map with Layer Controls

For datasets with multiple participants or groups, we can use layer controls to toggle different groups on and off:

```{r}
# artificially split data into two "participants" to demonstrate layer controls
gps_participants <- gps |>
  mutate(
    start_time = as.POSIXct(start),
    # split by time - early vs late attendees
    participant = ifelse(start_time < median(start_time), "Early Attendee", "Late Attendee")
  ) |>
  arrange(start_time)

# create base map
map <- leaflet() |>
  addTiles() |>
  setView(lng = mean(gps_participants$lon), lat = mean(gps_participants$lat), zoom = 13)

# define a function
add_participant_markers <- function(participant_name, map_obj, data, colors) {
  participant_data <- data |> filter(participant == participant_name)
  color_index <- which(unique(data$participant) == participant_name)
  
  map_obj |>
    addCircleMarkers(
      data = participant_data,
      lng = ~lon, 
      lat = ~lat,
      radius = 6,
      color = "#000",
      fillColor = colors[color_index],
      fillOpacity = 0.7,
      weight = 1,
      popup = ~paste0(
        "<strong>", name, "</strong><br>",
        "Group: ", participant, "<br>",
        "Time: ", format(start_time, "%H:%M"), "<br>",
        "Duration: ", round(duration/60, 1), " minutes"
      ),
      group = participant_name
    )
}

# get all participants in a list
participants <- unique(gps_participants$participant)
colors <- c("#C5050C", "#2E86AB")  # red, blue

# use reduce() to apply the function to each participant
map <- participants |>
  reduce(\(map_obj, participant) add_participant_markers(participant, map_obj, gps_participants, colors),
         .init = map)

# add layer controls to toggle groups
map |>
  addLayersControl(
    overlayGroups = unique(gps_participants$participant),
    options = layersControlOptions(collapsed = FALSE)
  )
```